---
output:
  reprex::reprex_document:
    venue: "gh"
    advertise: FALSE
    session_info: TRUE
    style: TRUE
    comment: "#;-)"
    tidyverse_quiet: FALSE
    std_out_err: TRUE
knit: reprex::reprex_render
---
---
title: "Workers' Compensation Loss-Cost EDA"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---



```{r load-data, echo=FALSE}

#install.packages("httpgd", dependencies = TRUE)


# Load necessary libraries
library(tidyverse)
library(lubridate)
library(ggplot2)

```


```{r load-data, echo=FALSE}
# Load dataset (Modify path if loading from a CSV)
train_df <- read.csv("train.csv", stringsAsFactors = FALSE)
test_df <- read.csv("test.csv", stringsAsFactors = FALSE)


# Convert dates to Date format
train_df$DateTimeOfAccident <- as.Date(train_df$DateTimeOfAccident)
test_df$DateTimeOfAccident <- as.Date(test_df$DateTimeOfAccident)

train_df$DateReported <- as.Date(train_df$DateReported)
test_df$DateReported <- as.Date(test_df$DateReported)
```     

```{r, ECHO = FALSE}

cat("Number of Rows in test_df:", nrow(test_df), "\n")
colnames(test_df)


```


There needs to be a Earned Premium column 

```{r create-synthetic-data, echo=FALSE}
set.seed(42)  # Ensure reproducibility
library(dplyr)

#Standardize Column Names
if ("InitialIncurredCalimsCost" %in% colnames(test_df)) {
  test_df <- test_df %>%
    rename(UltimateIncurredClaimCost = InitialIncurredCalimsCost)
}

#Function to Generate Earned Premiums & Loss Ratio
generate_synthetic_data <- function(df) {
  df <- df %>% filter(!is.na(WeeklyWages) & WeeklyWages > 0)

  # Generate synthetic Earned Premiums
  df$EarnedPremiums <- df$WeeklyWages * sample(seq(40, 60, by=5), nrow(df), replace=TRUE)

  # Compute Loss Ratio safely
  df$LossRatio <- df$UltimateIncurredClaimCost / df$EarnedPremiums

  return(df)
}

#Apply function to both train and test sets
train_df <- generate_synthetic_data(train_df)
test_df <- generate_synthetic_data(test_df)

#Print preview
head(train_df[, c("UltimateIncurredClaimCost", "EarnedPremiums", "LossRatio")])
head(test_df[, c("UltimateIncurredClaimCost", "EarnedPremiums", "LossRatio")])

```

```{r data creation check, echo = FALSE}
# Check if EarnedPremiums is missing or zero in test_df
cat("Missing EarnedPremiums in Test Set: ", sum(is.na(test_df$EarnedPremiums)), "\n")
cat("Zero EarnedPremiums in Test Set: ", sum(test_df$EarnedPremiums == 0, na.rm = TRUE), "\n")

# If EarnedPremiums is missing, re-run synthetic data generation
if (sum(is.na(test_df$EarnedPremiums)) > 0 | sum(test_df$EarnedPremiums == 0, na.rm = TRUE) > 0) {
  test_df <- generate_synthetic_data(test_df)
}

# Ensure LossRatio does not contain NA or Inf
test_df <- test_df %>%
  mutate(LossRatio = ifelse(is.finite(LossRatio), LossRatio, NA)) %>% 
  drop_na(LossRatio)

# Re-check missing values
cat("Final Missing LossRatio in Test Set: ", sum(is.na(test_df$LossRatio)), "\n")


```


```{r check loss ratio, ECHO = FALSE}

summary(train_df$LossRatio)
summary(test_df$LossRatio)

# Check if all values are zeros or very small
cat("Unique values in train_df$LossRatio:", length(unique(train_df$LossRatio)), "\n")
cat("Unique values in test_df$LossRatio:", length(unique(test_df$LossRatio)), "\n")

# Ensure LossRatio is numeric
train_df$LossRatio <- as.numeric(train_df$LossRatio)
test_df$LossRatio <- as.numeric(test_df$LossRatio)


```

There is a high right skew. 


```{r add-report-delay, echo=FALSE}
# Ensure both columns are in date format
train_df$DateTimeOfAccident <- as.Date(train_df$DateTimeOfAccident)
train_df$DateReported <- as.Date(train_df$DateReported)

test_df$DateTimeOfAccident <- as.Date(test_df$DateTimeOfAccident)
test_df$DateReported <- as.Date(test_df$DateReported)

# Compute Report Delay (Days between accident and report)
train_df$ReportDelay <- as.numeric(difftime(train_df$DateReported, train_df$DateTimeOfAccident, units="days"))
test_df$ReportDelay <- as.numeric(difftime(test_df$DateReported, test_df$DateTimeOfAccident, units="days"))

# Verify it’s added
colnames(train_df)
head(train_df$ReportDelay)
```

```{r create-accident-season, echo=FALSE}
train_df$AccidentSeason <- case_when(
  month(train_df$DateTimeOfAccident) %in% c(12, 1, 2) ~ "Winter",
  month(train_df$DateTimeOfAccident) %in% c(3, 4, 5)  ~ "Spring",
  month(train_df$DateTimeOfAccident) %in% c(6, 7, 8)  ~ "Summer",
  month(train_df$DateTimeOfAccident) %in% c(9, 10, 11) ~ "Fall",
  TRUE ~ NA_character_
)

test_df$AccidentSeason <- case_when(
  month(test_df$DateTimeOfAccident) %in% c(12, 1, 2) ~ "Winter",
  month(test_df$DateTimeOfAccident) %in% c(3, 4, 5)  ~ "Spring",
  month(test_df$DateTimeOfAccident) %in% c(6, 7, 8)  ~ "Summer",
  month(test_df$DateTimeOfAccident) %in% c(9, 10, 11) ~ "Fall",
  TRUE ~ NA_character_
)

# Convert to categorical (factor)
train_df$AccidentSeason <- as.factor(train_df$AccidentSeason)
test_df$AccidentSeason <- as.factor(test_df$AccidentSeason)

# Verify the column was created
table(train_df$AccidentSeason)
```

```{r check loss ratio zeros, echo = FALSE}

num_zeros <- sum(train_df$LossRatio == 0)
print(num_zeros)
````


To handle the right-skew of the Loss Ratio, I will apply a log transformation to compress the large values and spread out small values. Now the data will be more normally distributed. 

```{r check loss ratio distribution, echo = FALSE, fig.show="hold"}
# Check the distribution of Loss Ratio
train_df$LossRatio_log <- log1p(train_df$LossRatio)  # log(1 + LossRatio) to avoid log(0)
test_df$LossRatio_log <- log1p(test_df$LossRatio)

print(
  ggplot(train_df, aes(x = LossRatio_log)) +
    geom_histogram(bins = 30, fill = "blue", alpha = 0.6) +
    ggtitle("Log-Transformed Loss Ratio (Train Set)") +
    xlab("log(1 + Loss Ratio)") +
    ylab("Frequency") +
    theme_minimal()
)

print(
  ggplot(test_df, aes(x = LossRatio_log)) +
    geom_histogram(bins = 30, fill = "blue", alpha = 0.6) +
    ggtitle("Log-Transformed Loss Ratio (Test Set)") +
    xlab("log(1 + Loss Ratio)") +
    ylab("Frequency") +
    theme_minimal()
)



```


The distribution has a right skew, I will be using a tweedie regression. 

```{r check for missing values, echo = FALSE}

# Check for missing values
missing_values <- colSums(is.na(train_df))
missing_values


```

 There are no missing values 
```{r, echo =FALSE}

colnames(train_df)
str(train_df)  # Check structure of dataset


```

```{r, echo =FALSE}

install.packages("caret", type = "binary")
library(caret)  # Load caret
```


```{r, echo =FALSE}
dmy <- dummyVars("~ Gender", data = train_df)
train_df_dmy <- data.frame(predict(dmy, newdata = train_df))

# Print column names
print(colnames(train_df_dmy))
```

```{r, echo =FALSE}
colnames(train_df_dmy) <- c("Gender_Female", "Gender_Male", "Gender_Unknown")

train_df <- cbind(train_df, train_df_dmy)
str(train_df)  # Check structure of dataset

'Gender_Female' %in% names(train_df)
```

```{r, echo =FALSE}

dmy <- dummyVars("~ MaritalStatus", data = train_df)
train_df_dmy <- data.frame(predict(dmy, newdata = train_df))

# Print column names
print(colnames(train_df_dmy))

```
```{r, echo =FALSE}
colnames(train_df_dmy) <- gsub("^MaritalStatus", "Marital_", colnames(train_df_dmy))

# Merge with original dataset
train_df <- cbind(train_df, train_df_dmy)

# Check if all marital status categories exist
print(colnames(train_df))


sum(is.na(train_df$Marital_))  # Count NAs
sum(train_df$Marital_ == 0, na.rm = TRUE)  # Count empty strings


```

```{r, echo =FALSE}
head(train_df[c("Marital_", "Marital_M", "Marital_S", "Marital_U")])

sum(train_df$Marital_ == 0, na.rm = TRUE)  # Count empty strings

print(colnames(train_df))

train_df <- train_df[, !names(train_df) %in% c("Marital_")]


```

```{r, echo =FALSE}

dmy <- dummyVars("~ PartTimeFullTime", data = train_df)
train_df_dmy <- data.frame(predict(dmy, newdata = train_df))

# Print column names
print(colnames(train_df_dmy))
```

```{r, echo =FALSE}
colnames(train_df_dmy) <- c("FullTime", "PartTime")

# Merge with original dataset
train_df <- cbind(train_df, train_df_dmy)

print(colnames(train_df))

```

```{r, echo = FALSE}

# Encode Gender (use the same dummyVars model as train_df)
test_df_dmy <- data.frame(predict(dmy, newdata = test_df))  # Use the `dmy` object from before
colnames(test_df_dmy) <- c("Gender_Female", "Gender_Male", "Gender_Unknown")

# Merge into test_df
test_df <- cbind(test_df, test_df_dmy)

# Verify
print(colnames(test_df))

```

```{r, echo = FALSE}

# Apply dummy encoding for MaritalStatus (using the same method as train_df)
test_df_dmy <- data.frame(predict(dummyVars("~ MaritalStatus", data = train_df), newdata = test_df))

# Rename columns dynamically
colnames(test_df_dmy) <- gsub("^MaritalStatus", "Marital_", colnames(test_df_dmy))

# Merge into test_df
test_df <- cbind(test_df, test_df_dmy)

# Remove unwanted column if it appears
test_df <- test_df[, !names(test_df) %in% c("Marital_")]

# Check final structure
print(colnames(test_df))

```


```{r, echo = FALSE}

# Apply dummy encoding for PartTimeFullTime
test_df_dmy <- data.frame(predict(dummyVars("~ PartTimeFullTime", data = train_df), newdata = test_df))

# Rename properly
colnames(test_df_dmy) <- c("FullTime", "PartTime")

# Merge into test_df
test_df <- cbind(test_df, test_df_dmy)

test_df <- test_df[, !names(test_df) %in% c("PartTimeFullTimeF")]
test_df <- test_df[, !names(test_df) %in% c("PartTimeFullTimeP")]

# Check
print(colnames(test_df))


```

```{r, echo = FALSE}

print(setdiff(names(train_df), names(test_df)))  # Columns in train_df but not in test_df
print(setdiff(names(test_df), names(train_df)))  # Columns in test_df but not in train_df

train_df <- train_df[, !names(train_df) %in% c("ClaimDescription")]
test_df <- test_df[, !names(test_df) %in% c("ClaimDescription")]
"ClaimDescription" %in% names(train_df)  # Should return FALSE
"ClaimDescription" %in% names(test_df)   # Should return FALSE


```

Before fitting the model, visualize the relationships between LossRatio and predictors.

```{r, eco = FALSE}
install.packages("ggcorrplot")
library(ggcorrplot)

# Select only numerical columns
numeric_features <- train_df[, sapply(train_df, is.numeric)]

# Compute correlation matrix
cor_matrix <- cor(numeric_features, use = "complete.obs")

ggcorrplot(cor_matrix, method = "circle", type = "lower", lab = TRUE, show.legend = TRUE) +
  ggtitle("Correlation Matrix of Numerical Features") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels


# Find highly correlated features (threshold > 0.85)
high_corr_features <- findCorrelation(cor_matrix, cutoff = 0.85, names = TRUE)

# Print correlated features
print(high_corr_features)


```


```{r remove highly correlated feature, echo = FALSE}
train_df <- train_df[, !names(train_df) %in% c("WeeklyWages")]
test_df <- test_df[, !names(test_df) %in% c("WeeklyWages")]

"WeeklyWages" %in% names(train_df)  # Should return FALSE
"WeeklyWages" %in% names(test_df)   # Should return FALSE

```

Best Practice: Before running Boruta, If can remove highly correlated redundant features to make feature selection more robust.

```{r boruta, echo = FALSE}
install.packages("Boruta", type = "binary")
library(Boruta)
```


```{r boruta, echo = FALSE}
# Remove LossRatio and LossRatio_log from the dataset
train_df_no_target <- train_df[, !names(train_df) %in% c("LossRatio", "LossRatio_log")]


set.seed(42)
boruta_result <- Boruta(x = train_df_no_target, y = train_df$LossRatio_log, doTrace = 2)


```

```{r boruta results, echo = FALSE}
# Print summary of selected features
print(boruta_result)

plot(boruta_result, las = 2, cex.axis = 0.7)  

# Get the final selected features
final_features <- getSelectedAttributes(boruta_result, withTentative = FALSE)
```

```{r boruta results, echo = FALSE}
# Save to a CSV file
write.csv(final_features, "boruta_selected_features.csv", row.names = FALSE)

# Or save as an R object (faster to load later)
save(final_features, file = "boruta_features.RData")

```

```{r top 7 features, echo = FALSE}

# Extract feature importance scores from Boruta
importance_scores <- attStats(boruta_result)

# Sort features by importance
importance_scores <- importance_scores[order(-importance_scores$meanImp), ]

# Select the top 7 features
top_7_features <- rownames(importance_scores[1:7, ])
print(top_7_features)


```

When running Boruta for eature selection, these parameters help control verbosity(doTrace) and how I handle tentativee features(withTentative). doTrace controls how much information Boruta print while running. This helps with the runtime and understanding information about each feature at every step. Meanwhile, withTentative, helps determine whether I keep or discard the features that are classified under uncertain importance(Tentative). The best practices are to use doTrace =1 to monitor progress without excessive output. Also start with withTentative = False, but run TentativeRoughFix() if keeping more features is needed. 


To build a realistic loss-cost model, I must only use features available at the time of pricing or policy issuance.
Therefore, I will be removing the variables "UltimateIncurredClaimCost" and "InitialIncurredClaimCost".


# Load from CSV
final_features <- read.csv("boruta_selected_features.csv", stringsAsFactors = FALSE)$x

# OR load from RData file (recommended)
load("boruta_features.RData")

```{r, echo = FALSE}
top_7_features <- setdiff(top_7_features, c("UltimateIncurredClaimCost", "InitialIncurredCalimsCost"))
top_7_features <- setdiff(top_7_features, "Marital_U")

```

```{r, echo=FALSE}
print(top_7_features)
```
```{r split-data, echo=FALSE}
set.seed(42)  # Ensure reproducibility
library(caret)  # For data splitting

# Remove `LossRatio` and `MaritalStatus` from `train_df` and `test_df`
train_df <- train_df[, !names(train_df) %in% c("LossRatio", "MaritalStatus")]
test_df <- test_df[, !names(test_df) %in% c("LossRatio", "MaritalStatus")]

# Define 80% train, 20% validation split
train_index <- createDataPartition(train_df$LossRatio_log, p=0.8, list=FALSE)

# Create train and validation sets
train_set <- train_df[train_index, ]
validation_set <- train_df[-train_index, ]

# Remove `MaritalStatus` from `top_7_features`
top_7_features <- setdiff(top_7_features, "MaritalStatus")

# Now keep only the top 7 numeric features
train_df_top7 <- train_set[, c(top_7_features, "LossRatio_log")]
val_df_top7 <- validation_set[, c(top_7_features, "LossRatio_log")]
test_df_top7 <- test_df[, top_7_features]  # No target in test set

# Print dataset sizes
cat("Training Set Size: ", nrow(train_df_top7), "\n")
cat("Validation Set Size: ", nrow(val_df_top7), "\n")
cat("Test Set Size: ", nrow(test_df_top7), "\n")


```


```{r tweedie-model, echo=FALSE}
#install.packages("statmod")  # Run this once
install.packages("MASS")     # GLM models
install.packages("glmnet")   # Regularized models (optional)
library(statmod)
library(statmod)  # Required for Tweedie distribution
library(MASS)     # For GLM
library(glmnet)   # For potential Lasso/Ridge Tweedie models




```



```{r, echo=FALSE}

# Train Tweedie regression model
tweedie_model <- glm(LossRatio_log ~ ., 
                      data = train_df_top7, 
                      family = tweedie(var.power = 1.5, link.power = 0))  # 1 < var.power < 2 for insurance data

# Print model summary
summary(tweedie_model)


```


```{r, echo=FALSE}

library(car)
vif(tweedie_model)

library(ggplot2)

coef_df <- as.data.frame(coef(summary(tweedie_model)))
coef_df$Feature <- rownames(coef_df)
colnames(coef_df) <- c("Estimate", "StdError", "tValue", "pValue", "Feature")

ggplot(coef_df, aes(x = reorder(Feature, Estimate), y = Estimate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  ggtitle("Tweedie Model Feature Importance") +
  theme_minimal()


```

All VIF values are < 5, meaning NO severe multicollinearity!

```{r, echo=TRUE}

# Predict on validation set
val_predictions <- predict(tweedie_model, newdata = val_df_top7, type = "response")

# Compute RMSE
rmse <- sqrt(mean((val_predictions - val_df_top7$LossRatio_log)^2))
print(paste("Validation RMSE:", round(rmse, 4)))




```

Validation RMSE: 0.4662

```{r, echo = FALSE}

ggplot(data.frame(Predicted = val_predictions, Actual = val_df_top7$LossRatio_log), 
       aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +  
  theme_minimal() +
  ggtitle("Predicted vs Actual (Validation Data)")


```

The red line (ideal predictions) represents y = x (perfect predictions).
Most blue points (predictions) are below the red line, meaning the model often underestimates high loss ratios.
There is a strong clustering near zero, suggesting that the model struggles with high-loss cases.

It is best to check if the model is biased by plotting the residuals. 

```{r, echo = FALSE}
ggplot(data.frame(Residuals = val_predictions - val_df_top7$LossRatio_log), 
       aes(x = Residuals)) +
  geom_histogram(bins = 50, fill = "blue", alpha = 0.6) +
  theme_minimal() +
  ggtitle("Residuals Distribution")

```

The residuals are centered around zero meaning it is not baised. 
The residuals are heavily right-skewed.
There is a large peak near 0, meaning most predictions are close to the actual values.
There are some negative residuals, meaning the model over-predicts in certain cases.
The long tail on the left suggests some extreme errors (outliers).

The model is mostly predicting well, but some large negative residuals exist, meaning it occasionally overestimates losses.
The distribution is not perfectly normal, but that's expected in loss-cost modeling.
```{r, echo = FALSE}
ggplot(data.frame(Predicted = val_predictions, Residuals = val_predictions - val_df_top7$LossRatio_log), 
       aes(x = Predicted, y = Residuals)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  ggtitle("Residuals vs. Predictions")

```

The residuals fan out as predictions increase, meaning higher predicted loss ratios have larger errors.
There is a clear pattern where:
Low predictions have small residuals.
High predictions have larger, more spread-out residuals.

Heteroskedasticity is present: The variance of residuals increases with higher predicted values.
The model is not handling extreme losses well: It struggles to accurately predict high-loss cases, leading to high variance.


Log-Transform the variables 

```{r, echo = FALSE}
train_df_top7$LogEarnedPremiums <- log1p(train_df_top7$EarnedPremiums)
val_df_top7$LogEarnedPremiums <- log1p(val_df_top7$EarnedPremiums)
test_df_top7$LogEarnedPremiums <- log1p(test_df_top7$EarnedPremiums)



```

The model struggles with extreme cases.

```{r, echo = false}
library(statmod)

# Try different variance powers (1.1 to 1.9)
for (p in seq(1.1, 1.9, by = 0.2)) {
    model <- glm(LossRatio_log ~ ., 
                 data = train_df_top7, 
                 family = tweedie(var.power = p, link.power = 0))
    
    val_preds <- predict(model, val_df_top7, type = "response")
    rmse <- sqrt(mean((val_preds - val_df_top7$LossRatio_log)^2))
    
    print(paste("Tweedie Variance Power:", p, "Validation RMSE:", round(rmse, 4)))
}


```


```{r, echo = false}
tweedie_glm_interaction <- glm(LossRatio_log ~ EarnedPremiums * HoursWorkedPerWeek + Age, 
                               data = train_df_top7, 
                               family = tweedie(var.power = 1.1, link.power = 0))

summary(tweedie_glm_interaction)


```


```{r baseline, echo=false}

library(xgboost)

xgb_model <- xgboost(data = x_train, label = y_train, 
                     objective = "reg:tweedie", 
                     tweedie_variance_power = 1.1, 
                     nrounds = 200)

# Predict on validation set
xgb_predictions <- predict(xgb_model, x_val)

# Compute RMSE
rmse_xgb <- sqrt(mean((xgb_predictions - y_val)^2))
print(paste("XGBoost Tweedie Validation RMSE:", round(rmse_xgb, 4)))


```

XGBoost captures nonlinear relationships better than GLMs.
GLMs assume a specific distribution, but XGBoost can adapt better to real-world loss ratio patterns.
Regularization in GLM might not be sufficient, while XGBoost fine-tunes feature importance dynamically.

```{r, echo = FALSE}
library(xgboost)

# Convert data to XGBoost matrix format
dtrain <- xgb.DMatrix(data = x_train, label = y_train)
dval <- xgb.DMatrix(data = x_val, label = y_val)

# Set XGBoost parameters
params <- list(
  objective = "reg:tweedie",
  tweedie_variance_power = 1.1,
  eval_metric = "rmse",
  eta = 0.05,  # Learning rate
  max_depth = 5,
  subsample = 0.8,
  colsample_bytree = 0.8
)

# Train model with early stopping
xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 1000,  # High upper limit, but early stopping will prevent overfitting
  watchlist = list(val = dval),  # Validation set to monitor performance
  early_stopping_rounds = 50,  # Stops if no improvement in 50 rounds
  verbose = 1  # Shows training progress
)

# Get best number of rounds
best_nrounds <- xgb_model$best_iteration
print(paste("Best number of boosting rounds:", best_nrounds))



```


```{r, echo = FALSE}

# Predict using best iteration found by early stopping
xgb_predictions_tuned <- predict(xgb_model, dval, iteration_range = c(1, best_nrounds))

# Compute RMSE
rmse_xgb_tuned <- sqrt(mean((xgb_predictions_tuned - y_val)^2))
print(paste("Tuned XGBoost Tweedie Validation RMSE:", round(rmse_xgb_tuned, 4)))


```


```{r, echo = FALSE}

importance_matrix <- xgb.importance(model = xgb_model)
print(importance_matrix)

# Plot importance
xgb.plot.importance(importance_matrix)


```

```{r, echo = FALSE}

# Ensure test_df_top7 has only the top selected features
x_test <- as.matrix(test_df_top7[, top_7_features])

dtest <- xgb.DMatrix(data = x_test)

# Predict on test set using best iteration
test_predictions <- predict(xgb_model, dtest, iteration_range = c(1, 81))

# Convert back from log-transformed Loss Ratio
test_predictions <- expm1(test_predictions)  # Inverse of log1p()

# Store predictions
test_df_top7$Predicted_LossRatio <- test_predictions

# View final predictions
head(test_df_top7$Predicted_LossRatio)


```

