---
output:
  reprex::reprex_document:
    venue: "gh"
    advertise: FALSE
    session_info: TRUE
    style: TRUE
    comment: "#;-)"
    tidyverse_quiet: FALSE
    std_out_err: TRUE
knit: reprex::reprex_render
---
---
title: "Workers' Compensation Loss-Cost EDA"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---



```{r load-data, echo=FALSE}
# Load necessary libraries
library(tidyverse)
library(lubridate)
library(ggplot2)

```


```{r load-data, echo=FALSE}
# Load dataset (Modify path if loading from a CSV)
train_df <- read.csv("train.csv", stringsAsFactors = FALSE)
test_df <- read.csv("test.csv", stringsAsFactors = FALSE)


# Convert dates to Date format
train_df$DateTimeOfAccident <- as.Date(train_df$DateTimeOfAccident)
test_df$DateTimeOfAccident <- as.Date(test_df$DateTimeOfAccident)

train_df$DateReported <- as.Date(train_df$DateReported)
test_df$DateReported <- as.Date(test_df$DateReported)
```     

```{r, ECHO = FALSE}

cat("Number of Rows in test_df:", nrow(test_df), "\n")
colnames(test_df)


```


There needs to be a Earned Premium column 

```{r create-synthetic-data, echo=FALSE}
set.seed(42)  # Ensure reproducibility
library(dplyr)

#Standardize Column Names
if ("InitialIncurredCalimsCost" %in% colnames(test_df)) {
  test_df <- test_df %>%
    rename(UltimateIncurredClaimCost = InitialIncurredCalimsCost)
}

#Function to Generate Earned Premiums & Loss Ratio
generate_synthetic_data <- function(df) {
  df <- df %>% filter(!is.na(WeeklyWages) & WeeklyWages > 0)

  # Generate synthetic Earned Premiums
  df$EarnedPremiums <- df$WeeklyWages * sample(seq(40, 60, by=5), nrow(df), replace=TRUE)

  # Compute Loss Ratio safely
  df$LossRatio <- df$UltimateIncurredClaimCost / df$EarnedPremiums

  return(df)
}

#Apply function to both train and test sets
train_df <- generate_synthetic_data(train_df)
test_df <- generate_synthetic_data(test_df)

#Print preview
head(train_df[, c("UltimateIncurredClaimCost", "EarnedPremiums", "LossRatio")])
head(test_df[, c("UltimateIncurredClaimCost", "EarnedPremiums", "LossRatio")])

```

```{r data creation check, echo = FALSE}
# Check if EarnedPremiums is missing or zero in test_df
cat("Missing EarnedPremiums in Test Set: ", sum(is.na(test_df$EarnedPremiums)), "\n")
cat("Zero EarnedPremiums in Test Set: ", sum(test_df$EarnedPremiums == 0, na.rm = TRUE), "\n")

# If EarnedPremiums is missing, re-run synthetic data generation
if (sum(is.na(test_df$EarnedPremiums)) > 0 | sum(test_df$EarnedPremiums == 0, na.rm = TRUE) > 0) {
  test_df <- generate_synthetic_data(test_df)
}

# Ensure LossRatio does not contain NA or Inf
test_df <- test_df %>%
  mutate(LossRatio = ifelse(is.finite(LossRatio), LossRatio, NA)) %>% 
  drop_na(LossRatio)

# Re-check missing values
cat("Final Missing LossRatio in Test Set: ", sum(is.na(test_df$LossRatio)), "\n")


```


```{r check loss ratio, ECHO = FALSE}

summary(train_df$LossRatio)
summary(test_df$LossRatio)

# Check if all values are zeros or very small
cat("Unique values in train_df$LossRatio:", length(unique(train_df$LossRatio)), "\n")
cat("Unique values in test_df$LossRatio:", length(unique(test_df$LossRatio)), "\n")

# Ensure LossRatio is numeric
train_df$LossRatio <- as.numeric(train_df$LossRatio)
test_df$LossRatio <- as.numeric(test_df$LossRatio)


```


```{r check loss ratio distribution, echo = FALSE}

train_df$LossRatio_log <- log1p(train_df$LossRatio)  # log(1 + LossRatio) to avoid log(0)
test_df$LossRatio_log <- log1p(test_df$LossRatio)

ggplot(train_df, aes(x = LossRatio_log)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.6) +
  ggtitle("Log-Transformed Loss Ratio (Train Set)") +
  xlab("log(1 + Loss Ratio)") +
  ylab("Frequency") +
  theme_minimal()

ggplot(test_df, aes(x = LossRatio_log)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.6) +
  ggtitle("Log-Transformed Loss Ratio (Train Set)") +
  xlab("log(1 + Loss Ratio)") +
  ylab("Frequency") +
  theme_minimal()

```


The distribution has a right skew, I will be using a gamma regression. 

```{r check for missing values, echo = FALSE}

# Check for missing values
missing_values <- colSums(is.na(train_df))
missing_values


```

 There are no missing values 
```{r, echo =FALSE}

colnames(train_set)


```

```{r split-data, echo=FALSE}
set.seed(42)  # Ensure reproducibility
library(caret) # For data splitting

# Define 80% train, 20% validation split
train_index <- createDataPartition(train_df$LossRatio, p=0.8, list=FALSE)

# Create train and validation sets
train_set <- train_df[train_index, ]
validation_set <- train_df[-train_index, ]

# Print dataset sizes
cat("Training Set Size: ", nrow(train_set), "\n")
cat("Validation Set Size: ", nrow(validation_set), "\n")
cat("Test Set Size: ", nrow(test_df), "\n")


```


```{r install dependencies, echo=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("flexplot")


```

```{r gamma regression, echo = FALSE}


install.packages("devtools")  # Install devtools if not installed
devtools::install_github("dustinfife/flexplot")
library(flexplot)

```

install.packages("flexplot")  # Run this once if not installed
library(flexplot)

```

Before fitting the model, visualize the relationships between LossRatio and predictors.

```{r flexplot-univariate, echo=FALSE}

flexplot(LossRatio ~ 1, data=train_set)


```

```{r flexplot-bivariate, echo=FALSE}
# Bivariate plot: Loss Ratio vs. Weekly Wages
flexplot(LossRatio ~ WeeklyWages, data=train_set, method="Gamma", jitter=c(0, 0.2))

```

```{r gamma-models, echo=FALSE}
# Full Gamma Model
full_gamma <- glm(LossRatio ~ WeeklyWages + ReportDelay + AccidentSeason + MaritalStatus + Gender, 
                  data=train_set, family=Gamma(link="log"))

# Reduced Gamma Model (removing non-significant predictors)
reduced_gamma <- glm(LossRatio ~ WeeklyWages + ReportDelay, 
                     data=train_set, family=Gamma(link="log"))
```

